---
title: "Starting Analysis & Visualization of Spatial Data with R"
author: "Aaron Brown"
date: "Sunday, May 17, 2015"
output:
  pdf_document:
    highlight: tango
    number_sections: yes
    toc: yes
  html_document:
    number_sections: yes
    theme: united
    toc: yes
---

#Introduction
This introduction was taken as part of the [Spatial.ly: R Spatial Tips](http://spatial.ly/r/) site.  The site was built to give R users an introduction into spatial analysis using R.

This document is based on the workbooks discussed [here](http://spatialanalysis.co.uk/2013/04/analysis-visualisation-spatial-data/).
##Packages Required
This analysis uses several packages in the analysis.

  1. **ProjectTemplate** library to manage the files and structure of the project.
  2. **maptools** and **rgdal** as required on [this page](http://spatialanalysis.co.uk/2011/01/handling-spatial-data-with-r/).
  
This project is loaded by executing the following codechunk.
```{r}
library(knitr)
opts_knit$set(root.dir = '..')
```
```{r}
library(ProjectTemplate)
load.project()
```
##Files downloaded
The [London Sports Participate shapefile](http://spatialanalysis.co.uk/wp-content/uploads/2010/09/London_Sport1.zip) and [London Cycle Hire Locations](http://dl.dropbox.com/u/10640416/London_cycle_hire_locs.csv) are downloaded to the *./data* directory.
```{r}
if(!file.exists('./data/London_Sport1.zip')){
      url <- "http://spatialanalysis.co.uk/wp-content/uploads/2010/09/London_Sport1.zip"
      download(url, dest = "data/London_Sport1.zip", mode = "wb")
}
unzip("data/London_Sport1.zip", overwrite = TRUE, exdir = "data")

if(!file.exists('./data/London_cycle_hire_locs.csv')){
      url <- "http://dl.dropbox.com/u/10640416/London_cycle_hire_locs.csv"
      download(url, dest = "data/London_cycle_hire_locs.csv", mode = "wb")
}
```